\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}

### 2021-10-15 Manuscripts/Manuscript by chapters- SIM Revisions/03-GAM Theory.Rmd
Working on reviewers' comments

Splines (cubic, thin plate, etc.) are commonly used _basis functions_; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data [@wood2017;@simpson2018]. Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983]. Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. The timeline can be divided in equally spaced _knots_, each knot being a region where a different set of basis functions will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is set using default values in the package _mgcv_ depending on the number of knots. In Figure \@ref(fig:basis-plot)A, the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the "wiggliness" of the fit, each of the basis functions of Figure \@ref(fig:basis-plot)A is weighted by multiplying it by a coefficient according to the matrix of Figure \@ref(fig:basis-plot)B. The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the "wiggliness" of the smooth fit to prevent overfitting. A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

@@

   #GAM model of time, 5 knots
gm<-gam(y~s(time,k=5),data=dat)

@@

#length of dataframe to be created: number of knots by number of timepoints (minus 1 for the intercept that we won't plot)
ln<-6*(length(coef(gm)))

@@

To get the weighted basis functions, each basis (from Figure Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions. Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1). On the other hand, the third basis function has roughly doubled its value. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (orange line), along the simulated values per group, which appear as points.

(ref:basis-plot-caption) Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. B: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Weighted basis functions. Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.

>>>>

Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as _basis functions_, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness [@wegman1983]. Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.

Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness. The use of CS as smoothers in GAMs was discussed within the original GAM framework [@hastie1987], but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as 'knots', the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit [@wood2003]. In contrast, Wood [@wood2003] provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs. \normalfont{[@wood2003]} and [@wood2017]).

**To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable. Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).**

If five basis functions are used to fit a GAM for the data that appears in Figure \@ref(fig:l-q-response)C, the resulting fitting process is shown in Figure \@ref(fig:basis-plot)A. The four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the "wiggliness" of the fit, each of the basis functions of Figure \@ref(fig:basis-plot)A is weighted by multiplying it by a coefficient according to the matrix of Figure \@ref(fig:basis-plot)B. The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the "wiggliness" of the smooth fit to prevent overfitting. A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

@@

   #GAM model of time, 5 basis functions
gm<-gam(y~s(time,k=5),data=dat)

@@

#length of dataframe to be created: number of basis by number of timepoints (minus 1 for the intercept that we won't plot)
ln<-6*(length(coef(gm)))

@@

To get the weighted basis functions, each basis (from Figure Figure \@ref(fig:basis-plot)A) is multiplied by the corresponding coefficients in Figure \@ref(fig:basis-plot)B, thereby increasing or decreasing the original basis functions. Figure \@ref(fig:basis-plot)C shows the resulting weighted basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is **negative** and less than 1). On the other hand, the third basis function has roughly doubled its value. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of _time_ is shown in Figure \@ref(fig:basis-plot)D (orange line), along the simulated values per group, which appear as points.

(ref:basis-plot-caption) Basis functions for a single smoother for time. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. B: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Weighted basis functions. Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.
